---
title: "MoveTraits.V0.1 Spatial Annotation Tutorial"
author: "Anne G. Hertel"
date: "3/5/2026"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_float:
      collapsed: true
    code_folding: show
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, warning=FALSE,message=FALSE}
library(tidyverse);library(bit64); library(geodata);library(dggridR);library(terra); library(sf);library(terra); library(mapview);library(RColorBrewer)
```

# Download the database

Please download the three levels of the database from figshare and store in your local directory:

https://doi.org/10.6084/m9.figshare.28611890.v2

# Load MoveTraits database

The MoveTraits database version 0.1 has been compiled from open access data from Tucker et al. 2023 "Behavioral responses of terrestrial mammals to COVID-19 lockdowns" (https://zenodo.org/records/7704108, file "Tucker_Road_Spatial.rds"; only data from 2019) and from all openly available GPS datasets in Movebank (effective 03.2025). 

The database has three levels: species level trait summaries, individual level traits summaries, within individual repeated trait measures. The three levels are stored in separate files.

# Spatial grid
We have annotated traits with spatial grids to facilitate spatial annotation while protecting the underlying gps coordinates. We are using hexagonal equal area grids at approxiamately 1square kilometer (1km resolution), 100 square kilometer (10km resolution) and 10000 square kilometer (100km resolution). 

```{r}
dggs.100    <- dgconstruct(projection = "ISEA", area = 10000, resround='nearest')
dggs.10     <- dgconstruct(projection = "ISEA", area = 100, resround='nearest')
dggs.1      <- dgconstruct(projection = "ISEA", area = 1, resround='nearest')
```

An example how the grid looks like - we are here showing exemplary the grid cells at 100km resolution (10000 square km) intersecting Germany. 
```{r}
grid <- dgearthgrid(dggs.100, savegrid = NA, return_sf = TRUE) 

ger_sf <- map_data("world") %>%
  filter(region == "Germany") |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326, agr = "constant") %>%
  group_by(group) %>%
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON") %>%
  st_union() %>%
  st_sf() 

cropped_poly <- st_intersection(grid, ger_sf)

ggplot() +
  geom_sf     (data=ger_sf, color=alpha("black"), fill = NA) +
  geom_sf     (data=cropped_poly, color=alpha("black"), fill = NA) +
  theme_minimal() 

```

## Species level - not yet implemented

```{r}
#MoveTrait.v0.1.sp <- readRDS("MoveTrait.v0.1_species.sum_20251011.rds")

#dim(MoveTrait.v0.1.sp)
```


## Individual level

```{r}
MoveTrait.v0.1.ind <- readRDS("DATA/5.MB_indv_traitsum/10596067_10666985.rds")
```

The individual level database follows the same structure as the species level database. But contains more individual level information:

```{r, echo = F}
column_info <- data.frame(
  ColumnName = 
    c(names(MoveTrait.v0.1.ind)[1:6]
      ),
  Explanation = c(
    "Movebank study identifier",
    "Movebank individual identifier",
    "The individuals mean track longitude",
    "The individuals mean track latitude",
    "Intersecting 100km grid cell identities",
    "Intersecting 10km grid cell identities"
  )
)

knitr::kable(column_info, caption = "Column descriptions")
```

From cell identity to spatial annotation
```{r,echo=F}
grid.id.100km <- as.numeric(unlist(strsplit(MoveTrait.v0.1.ind$grid.id.100km, ";")))
print(grid.id.100km)
length(grid.id.100km)
```

The animal moved over 87 grid cells of 100km resolution (10000 square km). Therefore the trait summary is associated to that space. 

Where did the animal occur?
```{r}
intersect_cells_100 <- dgcellstogrid(dggs.100, grid.id.100km)
mapview(intersect_cells_100, map.types = c("OpenStreetMap.DE"), col.regions = "green", legend = FALSE)
```

Get the centroid coordinates of these cells to intersect with environmental layers
```{r}
lon <- dgSEQNUM_to_GEO(dggs.100, grid.id.100km)$lon_deg
lat <- dgSEQNUM_to_GEO(dggs.100, grid.id.100km)$lat_deg
centroids <- data.frame(grid.id.100km,lon,lat)
head(centroids)
```

Example - load and extract HFI values associated with the animals occurrence
```{r}
hfi <- geodata::footprint(year = 2009, path = ".")
plot(hfi)
```

```{r}
pts <- vect(centroids, crs = "EPSG:4326")
values <- terra::extract(hfi, pts, ID = FALSE, bind = TRUE)
head(values)
```

```{r}
values_sf <- sf::st_as_sf(values)
intersect_cells_100 <- intersect_cells_100 |> 
  left_join(st_drop_geometry(values_sf), by = c("seqnum" = "grid.id.100km")) 

mapview(intersect_cells_100, map.types = c("OpenStreetMap.DE"), 
        zcol = "wildareas.v3.2009.human.footprint",
        col.regions = colorRampPalette(rev(brewer.pal(11, "RdYlBu"))))
```

## Monthly summary

```{r}
MoveTrait.v0.1.month <- readRDS("DATA/6.MB_indv_monthly_traitsum/10596067_10666985.rds")
```

In the monthly summary we have summarized traits for each tracking month. These summaries can be spatially referenced using the grid.id.100km and grid.id.10km columns.

```{r, echo = F}
column_info <- data.frame(
  ColumnName = 
    c(names(MoveTrait.v0.1.month)[1:6]
      ),
  Explanation = c(
    "Movebank study identifier",
    "Movebank individual identifier",
    "Monitoring month",
    "Monitoring year",
    "Intersecting 100km grid cell identities",
    "Intersecting 10km grid cell identities"
  )
)

knitr::kable(column_info, caption = "Column descriptions")
```

The information, which cells were occupied in which months need to be unpacked from wide to long format, here exemplary done for grid.id.100km. The same procedure applies to the 10km grid.

```{r}
monthly_cells_100 <-
  MoveTrait.v0.1.month %>%
  separate_rows(grid.id.100km, sep = ";") %>%   
  mutate(grid.id.100km = as.numeric(grid.id.100km),
         month = str_pad(month, width = 2, pad = "0"),
         month_year = as.factor(paste(month,year,sep="_")))
```

The long format data column can then be used to identify which cells were occupied. 
```{r}
intersect_cells_100 <- dgcellstogrid(dggs.100, monthly_cells_100$grid.id.100km)
intersect_cells_100 <- 
  merge(intersect_cells_100,
        monthly_cells_100[,c("grid.id.100km","month_year")],
        by.x="seqnum",by.y="grid.id.100km")
```

We can now visualize where the individual had been in specific months.
```{r}
mapview(intersect_cells_100, map.types = c("OpenStreetMap.DE"), zcol = "month_year")
```

For range resident species, a gridded representation may be more useful.
```{r}
ggplot() +
    geom_sf(data = intersect_cells_100, fill = NA, colour = "black") +
    theme_classic() +
    facet_wrap(~month_year) +
    theme(axis.text = element_blank())
```

Trait summaries can be annotated with environmental conditions using time information (i.e. month and year) and monthly specific cell centroids. 

```{r}
lon_100km <- dgSEQNUM_to_GEO(dggs.100, monthly_cells_100$grid.id.100km)$lon_deg
lat_100km <- dgSEQNUM_to_GEO(dggs.100, monthly_cells_100$grid.id.100km)$lat_deg
centroids <- data.frame(monthly_cells_100,lon_100km,lat_100km)
head(centroids)
```

## Within Individual level

### How individual repeated trait measures will be annotated is still up for discussion!!

```{r}
MoveTrait.v0.1_rep.ind <- readRDS("DATA/7.MB_indv_trait/10596067_10666985.rds")
```

The third level of the database provides repeated individual trait measures. For each individual and every trait we provide the underlying repeated trait measures as a list of nested dataframes in columns 3 - 12

```{r}
names(MoveTrait.v0.1_rep.ind)
class(MoveTrait.v0.1_rep.ind[,5])
```

The underlying distribution for the trait "dmax24h" (maximum 24hr displacement) in column five, for the first individual in the database can be accessed in the following way. 

```{r}
MoveTrait.v0.1_rep.ind[,12][[1]]
```

For open access data, it contains the mean latitude and longitude of the particular day when the trait was measured. For closed access data it contains the identity of the grid cell that contains the mean coordinate, at either 10km, or 10 and 1km resolution, depending on the data owner's sharing selection. Only one coordinate is associated with every repeated trait measure and it is easy to link these back to the grid cell centroids.

```{r}
example.displ24h <-data.frame(MoveTrait.v0.1_rep.ind[,][[1]])
```

These centroids can be directly used to extract information from environmental layers.
```{r, echo = F}
pts <- terra::vect(example.displ24h, geom=c("lon.10km", "lat.10km"), crs = "EPSG:4326")
values <- terra::extract(hfi, pts, ID = FALSE, bind = TRUE)
```


```{r}
intersect_cells_10 <- dgcellstogrid(dggs.10, example.displ24h$grid.id.10km)
values_sf <- sf::st_as_sf(values)
intersect_cells_10 <- intersect_cells_10 |> 
  left_join(st_drop_geometry(values_sf), by = c("seqnum" = "grid.id.10km")) 

mapview(intersect_cells_10,
        zcol = "wildareas.v3.2009.human.footprint",
        col.regions = colorRampPalette(rev(brewer.pal(11, "RdYlBu"))))

```


